refnx - Neutron and X-ray reflectometry analysis in Python

### 2018-07-23 paper/manuscript.tex

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{geometry}
\usepackage[separate-uncertainty=true]{siunitx}
\geometry{legalpaper, portrait, margin=1in}
\usepackage[version=4]{mhchem}
\usepackage[affil-it]{authblk}

\begin{document}
\title{refnx - Neutron and X-ray reflectometry analysis in Python}
\author[1,*]{Andrew Nelson}
\author[2]{Stuart Prescott}

\affil[1]{% 
 Australian Nuclear Science and Technology Organisation, Locked Bag 2001, Kirrawee DC, NSW 2232, Australia}
\affil[2]{School of Chemical Engineering, University of New South Wales,  Sydney, NSW, Australia}
\affil[*]{Corresponding author: Andrew.Nelson@ansto.gov.au}

@@

@@

\section*{Introduction}
The use of specular X-ray and neutron reflectometry for morphological characterisation of thin films on the approximate size range 10 to \SI{5000}{\angstrom} has exploded over the past years \cite{Wood2017, Daillant2009}. Most neutron and X-ray sources have instruments to perform reflectometry measurements, and there is an ongoing need for accessible software programs for users of those instruments to analyse their data in a straightforward fashion, including the co-refinement of multiple contrast datasets.  Several programs are available for this purpose, with a variety of different features \cite{Bjorck2007, Gerelli2016, Kienzle2011, Nelson2006}. These programs typically create a model of the interface, and incrementally refine the model against the data using least-squares, or use Bayesian approaches \cite{Sivia2006, 2010arXiv1008.4686H} to examine the posterior probability distribution of the parameters (i.e. the statistical variation of the parameters in a model).

@@

\section*{Method}
\refnx\ is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter\ object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing the prior. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ can be applied using constraints.

@@

The \Objective\ statistics are used directly by the \textbf{CurveFitter} class to perform least-square fitting or Bayesian MCMC of the system. \refnx\ uses the \SciPy\ package for least-squares analysis (DifferentialEvolution, Levenberg Marquardt, LBFGSB - Limited Broyden--Fletcher--Goldfarb--Shanno with bounds) and the \emcee\ and \ptemcee\ packages \cite{emcee, ptemcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. \ptemcee\ is a fork of \emcee\ that implements parallel tempering, for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher temperatures, while individual modes are efficiently explored by chains at lower temperatures.
The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\ overriding the log-likelihood method.
Parallelisation of the sampling is done automatically, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is done using the \corner\ package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter\ value is updated and given a standard error. For the sampling these represent the median and half the [15.87, 84.13] percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

\begin{figure}
  \includegraphics[width=\linewidth]{components}
  \caption{Schematic showing the relationship between classes that make up a typical reflectometry curve-fitting problem.}
  \label{fig:components}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{./supporting_information/gui.png}
  \caption{Screenshot of the \Jupyter\/\ipywidgets\ GUI. \Jupyter\ notebook available in supporting information.}
  \label{fig:gui}
\end{figure}

\subsection*{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach.

@@

\begin{figure}%
\centering
\subfigure{%
\label{fig:global_fit}%
\includegraphics[width=5in]{./supporting_information/global_fit.png}}\hspace{1em}%
\subfigure{%
\label{fig:d2o_sld_spread}%
\includegraphics[width=5in]{./supporting_information/d2o_sld_spread.png}}\\%
\caption{a) Reflectivity from a DMPC bilayer measured at three contrasts, with median in red, and 500 samples from the posterior distribution. Dataset for \ce{Si} match offset by 0.1, \ce{H_2O} contrast offset by 0.001. b) SLD profile of the \ce{D_2O} model showing 500 samples from the posterior distribution, as well as the median.}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{./supporting_information/corner}
  \caption{Corner plot for the varying parameters of DMPC bilayers measured at three contrasts. The sampling took $\sim$13 minutes on a 3.2GHz quad-core computer for 30 saved steps, corresponding to 6000 samples, with the steps being thinned by a factor of 100.}
  \label{fig:corner}
\end{figure}

\section*{Distribution and Modification}
Each submodule in \refnx\ possesses its own unit testing code for checking that the functions and classes in the module operate correctly, both individually and collectively. For example, there are tests that check that the reflectivity of a model is calculated correctly, or that the behaviour of a function is correct for the different possible inputs and code paths through it. Since the test suite is an integral part of the package each installation is testable. In addition, there is a benchmarking suite to track changes in performance, specifically the speed of critical calculations, over time.

@@

\section*{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx\ in a \Jupyter\ notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx, \SciPy, \NumPy, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda\ package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda\ can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter\ notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx\ is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx\ is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

\section*{Conclusions}\label{conclusions}
\refnx\ is a powerful tool for least-squares or Bayesian analysis of neutron and X-ray reflectometry data that is ideally usable for reproducible research with \Jupyter\ notebooks, and with extensibility in mind. Its features include: MCMC sampling of posterior distribution for parameters, structural models constructed from modular components with physically relevant parameterisation, algebraic inter-parameter constraints, mixed area models, co-refinement of multiple datasets, probability distributions for parameter bounds used directly for log-prior terms, and a (\Jupyter\ ) \ipywidgets\ GUI.

\section*{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on \refnx\ development.

\subsection*{Supporting information}
\textbf{gui.ipynb} - \Jupyter\ notebook used to create the GUI screenshot.\\
\textbf{lipid.ipynb} - \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{lipid.pdf} - PDF view of the \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{refnx-paper.yml} - \conda\ environment file to reproduce the analysis environment in this paper.

\bibliography{main}{}
\bibliographystyle{unsrt}
\end{document}

>>>>

\documentclass[pdf,preprint]{iucr}
\RequirePackage{graphicx}
\journalcode{A}

% \usepackage{subfigure}
\usepackage{amsmath}
\usepackage[separate-uncertainty=true]{siunitx}
\usepackage[version=4]{mhchem}

\begin{document}
\title{refnx - Neutron and X-ray reflectometry analysis in Python}
\cauthor[a]{Andrew}{Nelson}{andrew.nelson@ansto.gov.au}{}
\author[b]{Stuart~W.}{Prescott}

\aff[a]{Australian Nuclear Science and Technology Organisation, Locked Bag 2001, Kirrawee DC, NSW 2232 \country{Australia}}
\aff[b]{School of Chemical Engineering, University of New South Wales,  Sydney, NSW, 2052 \country{Australia}}

@@

\begin{synopsis}
%TODO synopsis for ToC
\end{synopsis}

@@

\section{Introduction}
The use of specular X-ray and neutron reflectometry for morphological characterisation of thin films on the approximate size range 10 to \SI{5000}{\angstrom} has exploded over the past years \cite{Wood2017, Daillant2009}. Most neutron and X-ray sources have instruments to perform reflectometry measurements, and there is an ongoing need for accessible software programs for users of those instruments to analyse their data in a straightforward fashion, including the co-refinement of multiple contrast datasets.  Several programs are available for this purpose, with a variety of different features \cite{Bjorck2007, Gerelli2016, Kienzle2011, Nelson2006}. These programs typically create a model of the interface, and incrementally refine the model against the data using least-squares, or use Bayesian approaches \cite{Sivia2006, 2010arXiv1008.4686H} to examine the posterior probability distribution of the parameters (i.e. the statistical variation of the parameters in a model).

@@

\section{Method}
\refnx\ is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter\ object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing the prior. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ can be applied using constraints.

@@

The \Objective\ statistics are used directly by the \textbf{CurveFitter} class to perform least-square fitting or Bayesian MCMC of the system. \refnx\ uses the \SciPy\ package for least-squares analysis (DifferentialEvolution, Levenberg--Marquardt, LBFGSB - Limited Broyden--Fletcher--Goldfarb--Shanno with bounds) and the \emcee\ and \ptemcee\ packages \cite{emcee, ptemcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. \ptemcee\ is a fork of \emcee\ that implements parallel tempering, for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher temperatures, while individual modes are efficiently explored by chains at lower temperatures.
The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\ overriding the log-likelihood method.
Parallelisation of the sampling is done automatically, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is done using the \corner\ package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter\ value is updated and given a standard error. For the sampling these represent the median and half the [15.87, 84.13] percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

\begin{figure}
  \includegraphics[width=175mm]{components}
  \caption{Schematic showing the relationship between classes that make up a typical reflectometry curve-fitting problem.}
  \label{fig:components}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=85mm]{./supporting_information/gui.png}
  \caption{Screenshot of the \Jupyter/\ipywidgets\ GUI. \Jupyter\ notebook available in supporting information.}
  \label{fig:gui}
\end{figure}

\subsection{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach.

@@

\begin{figure}%
\centering
\label{fig:global_fit}%
\includegraphics[width=85mm]{./supporting_information/global_fit.png}%
\label{fig:d2o_sld_spread}%
\includegraphics[width=85mm]{./supporting_information/d2o_sld_spread.png}
\caption{a) Reflectivity from a DMPC bilayer measured at three contrasts, with median in red, and 500 samples from the posterior distribution. Dataset for \ce{Si} match offset by 0.1, \ce{H_2O} contrast offset by 0.001. b) SLD profile of the \ce{D_2O} model showing 500 samples from the posterior distribution, as well as the median.}
\end{figure}

\begin{figure}
  \includegraphics[width=175mm]{./supporting_information/corner.png}
  \caption{Corner plot for the varying parameters of DMPC bilayers measured at three contrasts. The sampling took $\sim$13 minutes on a 3.2GHz quad-core computer for 30 saved steps, corresponding to 6000 samples, with the steps being thinned by a factor of 100.}
  \label{fig:corner}
\end{figure}

\section{Distribution and Modification}
Each submodule in \refnx\ possesses its own unit testing code for checking that the functions and classes in the module operate correctly, both individually and collectively. For example, there are tests that check that the reflectivity of a model is calculated correctly, or that the behaviour of a function is correct for the different possible inputs and code paths through it. Since the test suite is an integral part of the package each installation is testable. In addition, there is a benchmarking suite to track changes in performance, specifically the speed of critical calculations, over time.

@@

\section{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx\ in a \Jupyter\ notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx, \SciPy, \NumPy, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda\ package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda\ can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter\ notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx\ is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx\ is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

\section{Conclusions}\label{conclusions}
\refnx\ is a powerful tool for least-squares or Bayesian analysis of neutron and X-ray reflectometry data that is ideally usable for reproducible research with \Jupyter\ notebooks, and with extensibility in mind. Its features include: MCMC sampling of posterior distribution for parameters, structural models constructed from modular components with physically relevant parameterisation, algebraic inter-parameter constraints, mixed area models, co-refinement of multiple datasets, probability distributions for parameter bounds used directly for log-prior terms, and a (\Jupyter\ ) \ipywidgets\ GUI.

\section{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on \refnx\ development.

\subsection{Supporting information}
\textbf{gui.ipynb} - \Jupyter\ notebook used to create the GUI screenshot.\\
\textbf{lipid.ipynb} - \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{lipid.pdf} - PDF view of the \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{refnx-paper.yml} - \conda\ environment file to reproduce the analysis environment in this paper.

\bibliography{main}{}
\bibliographystyle{iucr}
\end{document}
